Typed Functional Genetic Programming

### 2013-07-19 tex/thesis.tex
[tex]

@@

Let $V$ be infinite countable set of {\it 
variable names}. Let $C$ be set of {\it constant names}.	 	
Then $\Lambda$ is set of {\it \lterms} defined inductively as follows.	
\begin{align*}
x   \in V \cup C  &\then x     \in \Lambda \\
M,N \in \Lambda   &\then (M~N) \in \Lambda 
\textit{~~~~~~(Function application)} \\
x   \in V , M \in \Lambda &\then \lamb{x}{M} \in \Lambda
\textit{~~~~~($\lambda$-abstraction)} 
\end{align*}~
\end{definition}

@@

It is usual to use $V = \Sigma^+$ set of all non-empty finite strings of symbols 
from $\Sigma$ where $\Sigma$ is some alphabet set, e.g.  
$
\Sigma =
\setDots{a}{z}$.

@@

\begin{definition}
Let $A$ be set of {\it atomic type names}. \\
Then $\mathbb{T}$ is set of {\it types} inductively defined as follows.
\begin{align*}
\alpha      \in A  &\then   \alpha \in \T \\
\sigma,\tau \in \T &\then ( \sigma \ar  \tau ) \in \T 
\end{align*}~

@@

Usually A is called \textit{set of type variables}, 
but since it might be source of confusion with polymorphic type variables
we prefer this name, which also better suits our usage. 
\\

@@

% We will utilize this property ... asi netřeba řikat

\begin{todo}  
   & možná se hlobějc zmínit o curryingu  
\end{todo}

\section{Statement of a form $M : \sigma$}

	Let $\Lambda$ be set of {\it \lterms}. \\*
	Let $\mathbb{T}$ be set of {\it types}.       \\*
	A {\it statement} $M : \sigma$ is a pair $(M,\sigma) \in \Lambda \times \mathbb{T}$. \\*
	$M : \sigma$ is vocalized as {\it "$M$ has type $\sigma$"}.\footnote{ 
	$M : \sigma$ can be also imagined as $M \in \sigma$ } \\*
	The type $\sigma$ is the {\it predicate} and the term $M$ is the
	{\it subject} of the statement.

\section{Context}

	Let $\Gamma \in \mathfrak P \left({\Lambda \times  \mathbb{T}}\right)$. 
	($\Gamma$ is a set of {\it statements} of a form $M : \sigma$.)	\\*
	Then $\Gamma$ is {\it context} if it obeys following 
	conditions\footnote{
	The $\pi_1$ corresponds to the projection of the first component of the Cartesian product.
	}:
	\begin{align*}
		 \forall (x,\sigma) \in \Gamma &: x \in V \cup C \\
		 \forall s_1,s_2 \in \Gamma &: s_1 \neq s_2 \Rightarrow \pi_1(s_1) \neq \pi_1(s_2)
    \end{align*}

	In other words context is a set of statements with distinct variables or constants as subjects.

\begin{todo}
 & TALK ABOUT Context represents library/building blocks.
\end{todo}

\section{Statement of a form \GMS}

@@

	We construct valid statements of form \GMS by using inference rules.

\section{Inference rule}

@@

\section{Simply typed $\lambda$-calculus}

\begin{todo}
   & říct že se vždycky zastavěj (strong normalization) ale
     asi to říct na jinym místě než tady, tady se teprv definujou typy, 
     říct to někde kde už se ví jak sou napojený na termy    
\end{todo}

>>>>

\newcommand{\tur}[3]{#1 \vdash #2 : #3}

@@

Let $V$ be infinite countable set of {\it 
variable names}. Let $C$ be set of {\it constant names}, 
$V \cap C = \emptyset$.	 	
Then $\Lambda$ is set of {\it \lterms} defined inductively as follows.	
\begin{align*}
x   \in V \cup C  &\then x     \in \Lambda \\
M,N \in \Lambda   &\then (M~N) \in \Lambda 
\textit{~~~~~~(Function application)} \\
x   \in V , M \in \Lambda &\then \lamb{x}{M} \in \Lambda
\textit{~~~~~($\lambda$-abstraction)} 
\end{align*}~
\end{definition}

@@

It is usual to not distinguish between $V$ and $U$ and just use
one set of variable names. But because this distinction is quite convenient
for us in the implementation, we use it even here. 
Hopefully, it will not confuse the reader.

We can use $V = \Sigma^+$, set of all non-empty finite strings of symbols 
from $\Sigma$ where $\Sigma$ is some alphabet set, e.g.  
$
\Sigma =
\setDots{a}{z}$.

@@

\begin{definition}
Let $A$ be set of {\it atomic type names}. 
Then $\mathbb{T}$ is set of {\it types} inductively defined as follows.
\begin{align*}
\alpha      \in A  &\then   \alpha \in \T \\
\sigma,\tau \in \T &\then ( \sigma \ar  \tau ) \in \T 
\end{align*}~

@@

Usually A is called \textit{set of type variables}, 
but since it might be source of confusion with polymorphic type variables
we prefer this name, which also better suits our usage.

Specific definition of $A$ is not so important, but for practical purposes 
we can think of it as the set of all non-function types relevant in our
situation, e.g., $A = \{ Int , IntList , Bool, BoolList , Char, String \}$.
Or we can think of $A$ more generally as of set of all non-empty strings
of symbols of some alphabet. \\

@@

\subsection{Statements of the form $M : \sigma$}

\begin{definition}
	Let $\Lambda$ be set of {\it \lterms}. 
	Let $\mathbb{T}$ be set of {\it types}.       
	A {\it statement} $M : \sigma$ is a pair 
	$(M,\sigma) \in \Lambda \times \mathbb{T}$.
	\\ 
\end{definition}

Statement $M : \sigma$ is vocalized as {\it "$M$ has type $\sigma$"}.
The type $\sigma$ is called the {\it predicate} and 
the term $M$ is called the {\it subject} of the statement $M : \sigma$. \\

We can look on this relation in tree different ways.

\begin{description}
  \item[Programming point of view.] 
   Statement $M : \sigma$ means that value $M$ is 
   an instance of data type
   $\sigma$. So statement $M : \sigma$ is
   somehow analogous to \textit{Java}
   expression \texttt{$M$ instanceof $\sigma$}.

   Type \textit{constrains} the value 
   and gives \textit{information} about it, 
   e.g., function $f : \sigma \ar \tau$ 
   is constrained to only accept values of a type $\sigma$
   as input and must return value of a type $\tau$ as output.

  \item[Set point of view.] 
     Statement $M : \sigma$ can also be viewed as relation
     $M \in \sigma$. This makes us look on types as on
     collections of their instances.

  \item[Logic point of view.] According to 
  the \textit{Curry–Howard correspondence}
  we can read statement $M : \sigma$ as 
  \textit{"$M$ is a proof of formula $\sigma$"}.
  We will not treat this interesting subject in much detail,
  \lets just say that $\ar$ corresponds to the
  implication $\then$ and \textit{atomic types} correspond to
  \textit{propositional variables}. The whole system of
  \textit{simply typed $\lambda$-calculus} then corresponds
  to the \textit{implicational fragment of propositional 
  intuitionistic logic}. We are mentioning this beautiful
  correspondence in hope that it will help clarify 
  use of \textit{inference rules} which are key concept
  in the process of deriving valid statements about
  terms and types.

\end{description}

\subsection{Context}

%\begin{definition}
%Let $\Gamma \in \mathfrak P \left({\Lambda \times  \mathbb{T}}\right)$. 
%($\Gamma$ is a set of statements of a form $M : \sigma$.)	
%Then $\Gamma$ is {\it context} if it obeys following 
%conditions.
%%\footnote{The $\pi_1$ corresponds to the projection of the 
%%first component of the Cartesian product.}
%\begin{align*}
%\forall (x,\sigma) \in \Gamma &: x \in V \cup C  
%&\textit{(x is variable or constant)}\\
%\forall (x_1,\sigma_1),(x_2,\sigma_2) \in \Gamma &: 
%x_1,x_2 \in V , x_1 \neq x_2 \then \sigma_1 \neq \sigma_2
%&\textit{(variables are distinct)}
%\end{align*}~
%Ten druhej je navíc blbě, má tam bejt celej stejtment se nerovná pak...
%\end{definition}

\begin{definition}
~
\begin{enumerate}
 \item A \textit{declaration} is a statement 
 $x : \sigma$ where $x \in V \cup C$.

 \item A \textit{context} (or \textit{basis}) 
 is set of declarations with distinct variables as subjects.
\end{enumerate}~
\end{definition}

Again our definition slightly differs from the standard one
in that it permits more than one declaration with the same 
\textit{constant}, which enables us to simulate very primitive
kind of ad-hoc polymorphism by \textit{"constant overloading"}.  
And again it does not affect the theory; it just seems less
confusing to me to state these definitions as they are later 
used in the implementation.   \\

In our system, the set of \textit{building blocks} 
(the equivalent of the set $T \cup F$ in classical GP)
will be a context with all declarations having \textit{constants}
as subjects, i.e., a set of constants together with their types.

In the process of term generation, 
this initial context is being further extend with variables
 --- by "entering inner scopes of bodies of lambda functions".

%\subsection{Statements of the form \GMS}

%\subsection{Inference rules}
\subsection{Simply typed $\lambda$-calculus}

\begin{definition}
A statement $M : \sigma$ is \textit{derivable from}
a context $\Gamma$ if it can be produced by the following rules.
\begin{align*}
(x,\sigma) \in \Gamma &~\then~ \tur{\Gamma}{x}{\sigma}\\
\tur{\Gamma}{M}{(\sigma \ar \tau)}~,~\tur{\Gamma}{N}{\sigma} 
&~\then~ \tur{\Gamma}{(M~N)}{\tau}\\  
\tur{\Gamma \cup (x,\sigma)}{M}{\tau}
&~\then~ \tur{\Gamma}{\lamb{x}{M}}{(\sigma \ar \tau)} 
\end{align*}~
\end{definition}

@@

We construct valid statements of the form \GMS 
by using inference rules.\\

@@

\subsection{Simply typed $\lambda$-calculus}

\begin{todo}
   & tady to nějak okomentovat že to je teda vono
   & říct že se vždycky zastavěj (strong normalization) ale
     asi to říct na jinym místě než tady, tady se teprv definujou typy, 
     říct to někde kde už se ví jak sou napojený na termy    
\end{todo}
